---
title: "ELITE Portal Tutorial: Joining Metadata"
author: "Melissa Klein (Sage Bionetworks)"
date: "`r Sys.Date()`"
format: 
  html: 
    toc: true
    toc-depth: 3
    df-print: paged
knit: (function(input_file, encoding) {
   out_dir <- 'docs';
   rmarkdown::render(input_file,
     encoding=encoding,
     output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r, set-opts, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  print.rows = 10
)

# Prevent individual IDs from printing out in scientific notation
options(scipen = 999)
```

#### Install Synapser if you have not already

```{r install-synapser, eval = FALSE}
install.packages("synapser", repos = c("http://ran.synapse.org"))
install.packages(c("tidyverse", "lubridate"))
```

#### Load libraries

```{r load-libraries, message=FALSE, warning=TRUE}
library(synapser)
library(readr)
library(dplyr)
library(magrittr)
```

#### Log in to Synapse

```{r synlogin, eval=FALSE}
synLogin()
```

```{r synlogin_run, include = FALSE}
# For prettier rendered display: This executes the code without showing the
# printed welcome statements. The next block will show the code but not run it.
synLogin(authToken = "INSERT AUTH TOKEN HERE")
```

#### Download and read in metadata files

There are 3 metadata files that you will want to have access to and potentially join together to understand data you are looking at on the ELITE Portal: Individual, Biospecimen, and Assay.

This example utilizes the three metadata files that can be found in the [Study Details Page](https://eliteportal.synapse.org/Explore/Studies/DetailsPage/StudyData?studyKey=Mouse_M005_Study_Metabolomics) for the Mouse M005 Metabolomics Study.

The query below is generated when you download programmatically directly from the ELITE Portal.

```{r portal-query}
# Download the results of the filtered table query
query <- synTableQuery("SELECT * FROM syn52234677 WHERE ( ( \"Study\" = 'Mouse_M005_Study_Metabolomics' ) ) AND ( `resourceType` = 'metadata' )")
read.table(query$filepath, sep = ",")

# View the file path of the resulting csv
query$filepath
```

Now that the files have been downloaded, we can read them into R.

```{r read-metadata-files}
# Individual metadata
individual_metadata <- read_csv("files/individual_non_human_M005_Longevity Consortium_11-11-2024_final.csv", show_col_types = FALSE)

# Biospecimen metadata
biospecimen_metadata <- read_csv("files/biospecimen_non_human_M005_Longevity Consortium_11-11-2024_final.csv", show_col_types = FALSE)

# Assay metadata
assay_metadata <- read_csv("files/synapse_storage_manifest_assaymetabolomicstemplate.csv", show_col_types = FALSE)
```

#### Join Metadata

Now, we join the metadata files together using left joins, matching on specimenID, then on individualID.

```{r join-metadata}
# join all the rows in the assay metadata that have a match in the biospecimen metadata
joined_meta <- assay_metadata |> 
  
  #join rows from biospecimen that match specimenID
  left_join(biospecimen_metadata, by = "specimenID") |>
  
  # join rows from individual that match individualID
  left_join(individual_metadata, by = "individualID")

joined_meta
```

Congratulations! You have now bulk downloaded and joined metadata files!
